Global adjoint tomography -- Model GLAD-M25\footnote{This paper is dedicated to Dimitri Komatitsch, computational geophysicist extraordinaire, who passed away unexpectedly shortly before its submission.}

### 2019-02-07 main.tex
add reference

The deployment of USArray has dramatically increased data overage across the
continental United States.
In this section, we compare GLAD-M25 in this region 
with two global models, S362ANI$+$M and S40RTS, and two regional models,
US22~\citep{zhu2017radial} and SL2013NA~\citep{schaeffer2014imaging}.
US22 is a radially anisotropic model based on adjoint tomography using
USArray data from 180 regional earthquakes using 15--50~s
body waves and 25--150~s surface waves.
SL2013NA is a global upper-mantle shear wavespeed model based on multimode
surface waveform tomography primarily focused on North America.

North America is characterized by a large, high-wavespeed lithospheric craton,
bounded by the Rocky Mountain Front to the West and continental
margin to the East. On both sides the lithosphere has been deformed by
low wavespeed structures, thus forming sharp tectonic boundaries
~\citep{grand1984upper,masters1996shear,grand1997high,
megnin2000three,gu2001models,schaeffer2014imaging,zhu2017radial}.

At a depth of 100~km, we observe that the edges of the craton are sharper in
GLAD-M25 than in the other two global models,
and quite similar in shape and sharpness to the two regional models.
We can clearly see impressions of the Snake River Plain (with Yellowstone)
and the Raton hotspot, which also have distinct slow expressions at depths of 200~km.
The core of the craton exhibits fast wavespeeds at depths reaching 300~km Northwest of
Hudson Bay.
In the regional models the craton has basically faded away at these depths.
The Wyoming and Medicine Hat Blocks show a very strong and thick lithospheric
root from shallow depths to $\sim$400~km, similar to US22 but
difficult to discern in SL2013NA.
Similar observations can be made for the Yavapai and
Mazatzal Blocks, which define the southern boundary of the large craton.

@@

In the North, we see a high wavespeed craton beneath Greenland which is partly
separated from the North American craton by low wavespeed structures beneath
Baffin Bay and the Labrador Sea, where the North American craton nicely follows
the coastline.
In the Northeast corner of the maps we see that the Iceland hotspot is very
prominent in GLAD-M25 in all depth ranges, similar to S40RTS but largely missing
in S362ANI$+$M. In the Northwest we see a clear imprint of the Aleutian
subduction zone in the 200--400~km depth range, comparable to SL2013NA but absent
in the other global models.

@@

A defining tectonic feature of Europe is the Trans-European Suture Zone~(TESZ),
also known as the Tornquist-Teisseyre Zone,
separating the Precambrian East European Craton~(EEC) in the Northeast from
Phanerozoic Europe in the Southwest.
The TESZ is very consistent across all models,
even though is much sharper in GLAD-M25 and EU60.

At depths of 100---250~km we observe high wavespeeds beneath the Central Graben
in the North Sea, which is clearly separated from the EEC.
The tectonic history of the Mediterranean is complex,
involving several regional plates and multiple past and present subduction zones.
We see slab ponding in the transition zone beneath the western Mediterranean
associated with slab roll-back of the Appennines-Maghrebides subduction zone.
At shallower depths, we see low wavespeeds along its western border,
which are interpreted as back-arc extension associated with  roll-back.
Compared to S362ANI$+$M and S40RTS,
our model shows much sharper features for these tectonic structures.

@@

In Central Europe we note the Cenozoic Ridge System, characterized by low 
wavespeeds from the Massif Central to the Eifel. Both EU60 and
GLAD-M25 exhibit a slow anomaly beneath the northern part of the Rhine Graben,
which is interpreted as the reservoir that fuels the Eifel hotspot.

>>>>

The deployment of USArray has dramatically increased data coverage across the
continental United States.
In this section, we compare GLAD-M25 in this region 
with two global models, S362ANI$+$M and S40RTS, and two regional models,
US22~\citep{zhu2017radial} and SL2013NA~\citep{schaeffer2014imaging}.
US22 is a radially anisotropic model based on adjoint tomography using
USArray data from 180 regional earthquakes using 15--50~s
body waves and 25--150~s surface waves.
SL2013NA is a global upper-mantle shear wavespeed model based on multimode
surface waveform tomography primarily focused on North America.

North America is characterized by a large, high-wavespeed lithospheric craton,
bounded by the Rocky Mountain Front to the West and continental
margin to the East~\citep{grand1984upper, whitmeyer2007tectonic}.
On both sides the lithosphere has been deformed by
low wavespeed structures, thus forming sharp tectonic boundaries
~\citep{masters1996shear, grand1997high, megnin2000three, gu2001models}.

At a depth of 100~km, we observe that the edges of the craton are sharper in
GLAD-M25 than in the other two global models,
and quite similar in shape and sharpness to the two regional models.
We can clearly see impressions of the Snake River Plain (with Yellowstone)
and the Raton hotspot, which also have distinct slow expressions at depths of 200~km.
The core of the craton exhibits fast wavespeeds at depths reaching 300~km Northwest of
Hudson Bay.
In the regional models the craton has basically faded away at these depths.
The Wyoming and Medicine Hat Blocks show a very strong and thick lithospheric
root from shallow depths to $\sim$400~km, similar to US22 but
difficult to discern in SL2013NA.
Similar observations can be made for the Yavapai and
Mazatzal Blocks, which define the southern boundary of the large craton
~\citep{schaeffer2014imaging}.

@@

In the North, we see a high wavespeed craton beneath Greenland which is partly
separated from the North American craton by low wavespeed structures beneath
Baffin Bay and the Labrador Sea, where the North American craton nicely follows
the coastline~\citep{chalmers2001development}.
In the Northeast corner of the maps we see that the Iceland hotspot is very
prominent in GLAD-M25 in all depth ranges, similar to S40RTS but largely missing
in S362ANI$+$M. In the Northwest we see a clear imprint of the Aleutian
subduction zone in the 200--400~km depth range, comparable to SL2013NA but absent
in the other global models.

@@

A defining tectonic feature of Europe is the Trans-European Suture Zone~(TESZ),
also known as the Tornquist-Teisseyre Zone,
separating the Precambrian East European Craton~(EEC) in the Northeast from
Phanerozoic Europe in the Southwest~\citep{zielhuis1994deep}.
The TESZ is very consistent across all models,
even though is much sharper in GLAD-M25 and EU60.

At depths of 100---250~km we observe high wavespeeds beneath the Central Graben
in the North Sea, which is clearly separated from the EEC.
The tectonic history of the Mediterranean is complex,
involving several regional plates and multiple past and present subduction zones
~\citep{dewey1989kinematics}.
We see slab ponding in the transition zone beneath the western Mediterranean
associated with slab roll-back of the Appennines-Maghrebides subduction zone
~\citep{wortel2000subduction, zhu2012structure}.
At shallower depths, we see low wavespeeds along its western border,
which are interpreted as back-arc extension associated with  roll-back.
Compared to S362ANI$+$M and S40RTS,
our model shows much sharper features for these tectonic structures.

@@

In Central Europe we note the Cenozoic Ridge System~\citep{ziegler1992european},
characterized by low 
wavespeeds from the Massif Central to the Eifel. Both EU60 and
GLAD-M25 exhibit a slow anomaly beneath the northern part of the Rhine Graben,
which is interpreted as the reservoir that fuels the Eifel hotspot
~\citep{goes1999lower, zhu2015seismic}.
